---
title: 'MA40198: Lab sheet 6'
author: "Karim Anaya-Izquierdo"
output:
  html_document:
    number_sections: no
    toc: yes
    toc_float: yes
    code_download: yes
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight = TRUE)
library(tidyverse)
```

## Instructions

-   This assignment should be done in groups according to the coursework
    group allocation.

-   As indicated in each question, you should complete the answers by
    filling-out the 'Lab6.Rmd' file. You can remove the instructions and
    introductory sections if you wish.

-   The 'Lab6.Rmd' file can be downloaded by clicking in the **Code
    download** button at the top right of this page. If the button does
    not work you can download the 'Lab4.Rmd' file from the Moodle page.
    You should open this file in RStudio.

-   After completing the answers, each group should convert the
    'Lab6.Rmd' file into an HTML file by following the instructions in
    the <a href="#knit-to-html">Knit to html</a> section at the end of
    this file!

-   Each group should submit two files: the completed 'Lab6.Rmd' file
    and the created HTML file (which should be named 'Lab6.html') to the
    <a href="https://moodle.bath.ac.uk/mod/assign/view.php?id=1175047">submission
    point in Moodle</a>

-   This is lab is a **formative assessment** and you will receive
    feedback to your submitted work on an informal basis.

## Breaking strength of fibres {#fibre-strength-data}

A tensile strength experiment on single carbon fibres produced the
following data which gives the stress values (in giga-pascals) at which
each single $n=257$ carbon fibres failed. Fibres were of four different
lengths: 1 mm, 10 mm, 20 mm and 50 mm. The data is displayed in the
folowing table.

```{r,echo=FALSE}
fibres<-
  read.table(url("http://people.bath.ac.uk/kai21/ASI/data/strength.txt"),header = TRUE)
DT::datatable(
fibres,
  fillContainer = FALSE, options = list(pageLength = 20)
)
```

The fibre strength data can be read as follows:

```{r}
strength.data<-
  read.table(url("http://people.bath.ac.uk/kai21/ASI/data/strength.txt"),header = TRUE)

```

Mechanical Engineers are interested in establising a relationship
between length of fibres and the corresponding breaking tensile stress.
The so-called **weakest link hypothesis** establishes the notion that a
fibre consists of independent links, whose weakest member determines the
breaking of the fibre.

The idea behind this is that if $Y_1,Y_2$ are independent and
identically distributed random variables with common survival function
$S(y)$ then the survival function of the minimum $Y=\min(Y_1,Y_2)$ is
given by
$$P(Y>y)=P(\min(Y_1,Y_2)>y)=P(Y_1>y,Y_2>y)=P(Y_1>y)P(Y_2>y)=[S(y)]^2$$
Now let $S_L(y)$ be the probability that a fibre of length $L$ fails
under a tensile stress larger than $y$. The weakest link hypothesis can
be expressed as follows: \begin{equation}
S_{\scriptsize L_1+L_2}(y)=S_{\scriptsize L_1}(y)S_{\scriptsize L_2}(y)
\end{equation}

### Strength distribution modelling

One family of distributions that satisfies the weakest link hypothesis
is the Weibull family with shape parameter $\eta>0$ and scale parameter
$\tau\,L^{-1/\eta}$ where $\tau>0$. So in this case, the distribution of
the breaking tensile stress depends on the length $L$ of the fibre
through the scale parameter. The corresponding survival function for a
fibre of length $L$ is given by

$$S_L(y|\tau,\eta)=\exp\left(-\left(\frac{y}{\tau\,L^{-1/\eta}}\right)^{\eta}\right)=\exp\left(-L\,\left(\frac{y}{\tau}\right)^{\eta}\right)\,,\quad y>0$$
We will denote this model by $\mathcal F_0$. This model has $p=2$
unknown parameters.

Consider now the extended model which is a generalised Weibull
distribution with shape parameters $\eta>0$ and $\kappa>0$ and scale
parameter $\tau\,L^{-1/\eta}$ where $\tau>0$. The corresponding survival
function for a fibre of length $L$ is given by

$$S_L(y|\tau,\eta,\kappa)=\exp\left[-\frac{1}{\kappa}\left\{\left(1+\left(\frac{y}{\tau\,L^{-1/\eta}}\right)^\eta\right)^\kappa -1\right\}\right]=\exp\left[-\frac{1}{\kappa}\left\{\left(1+L\left(\frac{y}{\tau}\right)^\eta\right)^\kappa -1\right\}\right]\,,\quad y>0$$
Note this model only satisfy the weakest link hypothesis when
$\kappa=1$. We will denote this model by $\mathcal F_1$. This model has
$p=3$ unknown parameters

The final model we will consider is the one in which the breaking
tensile stress distribution for each length is unrelated to each other
and follows a generalised Weibull distribution. The corresponding
survival function for a fibre of length $L$ is given by

$$S_L(y|\tau_{\scriptsize L},\eta,\kappa)=\exp\left[-\frac{1}{\kappa}\left\{\left(1+\left(\frac{y}{\tau_{\scriptsize L}}\right)^{\eta}\right)^{\kappa} -1\right\}\right]\,,\quad y>0$$

Clearly, this model satisfies the weakest link hypothesis only when
$\kappa=1$ and $\tau_{\scriptsize L}=\tau\,L^{-1/\eta}$ for some unknown
$\tau>0$. We will denote this model by $\mathcal F_2$. This model has
$p=6$ unknown parameters since $\tau_{\scriptsize 1}$,
$\tau_{\scriptsize 10}$, $\tau_{\scriptsize 20}$ and
$\tau_{\scriptsize 50}$ are all unknown.

## Question 1

Assuming that the breaking stress corresponds to an independent data
from an unknown member of $\mathcal F_i$ for $i=0,1,2$. In each case,
estimate the unknown generating parameters by maximum likelihood. To
perform the optimisation, you should use the following
reparametrisations:

$$
\boldsymbol{\theta}=
\left(
\begin{array}{c}
\theta_1\\
\theta_2
\end{array}
\right)
=
\left(
\begin{array}{c}
\log \eta\\
\log \tau
\end{array}
\right) \qquad \mbox{for } \mathcal F_0
$$

$$
\boldsymbol{\theta}=
\left(
\begin{array}{c}
\theta_1\\
\theta_2 \\
\theta_3
\end{array}
\right)
=
\left(
\begin{array}{c}
\log \eta\\
\log \tau \\
\log \kappa
\end{array}
\right)  \qquad \mbox{for } \mathcal F_1
$$

$$
\boldsymbol{\theta}=
\left(
\begin{array}{c}
\theta_1\\
\theta_2 \\
\theta_3\\
\theta_4 \\
\theta_5\\
\theta_6 
\end{array}
\right)
=
\left(
\begin{array}{c}
\log \eta\\
\log \tau_{\scriptsize 1}\\
\log \tau_{\scriptsize 10}\\
\log \tau_{\scriptsize 20}\\
\log \tau_{\scriptsize 50}\\
\log \kappa
\end{array}
\right)  \qquad \mbox{for } \mathcal F_2
$$

```{r}

nll_F0 <- function(theta, data){
  l = data[2]
  y = data[1]
  eta = exp(theta[1])
  tau = exp(theta[2])
  nll = -sum(-l*(y/tau)^eta)
  return(nll)
}

nll_F1 <- function(theta, data){
  l = data[2]
  y = data[1]
  eta = exp(theta[1])
  tau = exp(theta[2])
  kap = exp(theta[3])
  nll = -sum(-(1/kap)*((1+l*(y/tau)^eta)^kap-1))
  return(nll)
}

nll_F2 <- function(theta,data){
  l = data[2]
  y = data[1]
  eta = exp(theta[1])
  tau_1 = exp(theta[2])
  tau_10 = exp(theta[3])
  tau_20 = exp(theta[4])
  tau_50 = exp(theta[5])
  kap = exp(theta[6])
  nll_1 = nll_F1(c(theta[1],theta[2],theta[6]), data %>% filter(length==1))
  nll_10 = nll_F1(c(theta[1],theta[3],theta[6]), data %>% filter(length==10))
  nll_20 = nll_F1(c(theta[1],theta[4],theta[6]), data %>% filter(length==20))
  nll_50 = nll_F1(c(theta[1],theta[5],theta[6]), data %>% filter(length==50))
  nll = nll_1 + nll_10 + nll_20 + nll_50
  return(nll)
}

fit_F0 <- optim(c(0,0), nll_F0, method = 'BFGS', data = strength.data)
fit_F1 <- optim(c(0,0,0), nll_F1, method = 'BFGS', data = strength.data)
fit_F2 <- optim(c(0,0,0,0,0,0), nll_F2, method = 'BFGS', data = strength.data)
fit_F0$par
fit_F1$par
fit_F2$par

```

## Question 2

Consider now the model $\mathcal F_3$ a Weibull family with shape
parameter $\eta_{\scriptsize L}=\eta\,L^\nu$ and scale parameter
$\tau\,L^{-1/\eta_{\scriptsize L}}$ so that both shape and scale depend
on the length of the fibres and $\nu\in R$. The survival is given by
$$S_L(y|\eta,\tau,\nu)=\exp\left(-\left(\frac{y}{\tau\,L^{-1/\eta_{\scriptsize L}}}\right)^{\eta_{\scriptsize L}}\right)=\exp\left(-L\,\left(\frac{y}{\tau}\right)^{\eta\,L^\kappa}\right)\,,\quad y>0$$
This model has $p=3$ unknown parameters.

Finally, consider the saturated model $\mathcal F_4$ of $p=12$ unknown
parameters where the distribution of the breaking stress is generalised
Weibull with different parameters $\eta$, $\tau$ and $\kappa$ for each
length.

Obtain the corresponding MLE's for $\mathcal F_3$ and $\mathcal F_4$.
For the optimisation use

$$
\boldsymbol{\theta}=
\left(
\begin{array}{c}
\theta_1\\
\theta_2 \\
\theta_3
\end{array}
\right)
=
\left(
\begin{array}{c}
\log \eta\\
\log \tau \\
\nu
\end{array}
\right)  \qquad \mbox{for } \mathcal F_3
$$ and use the logarithm of all parameters for $\mathcal F_4$.

```{r}
nll_F3 <- function(theta, data){
  l = data[2]
  y = data[1]
  eta = exp(theta[1])
  tau = exp(theta[2])
  xi = theta[3]
  nll = -sum(-l*(y/tau)^(eta*l^xi))
  return(nll)
}

nll_F4 <- function(theta, data){
  l = data[2]
  y = data[1]
  eta_1 = exp(theta[1])
  eta_2 = exp(theta[2])
  eta_3 = exp(theta[3])
  eta_4 = exp(theta[4])
  tau_1 = exp(theta[5])
  tau_10 = exp(theta[6])
  tau_20 = exp(theta[7])
  tau_50 = exp(theta[8])
  kap_1 = exp(theta[9])
  kap_10 = exp(theta[10])
  kap_20 = exp(theta[11])
  kap_50 = exp(theta[12])
  nll_1 = nll_F1(c(theta[1],theta[5],theta[9]), data %>% filter(length==1))
  nll_10 = nll_F1(c(theta[2],theta[6],theta[10]), data %>% filter(length==10))
  nll_20 = nll_F1(c(theta[3],theta[7],theta[11]), data %>% filter(length==20))
  nll_50 = nll_F1(c(theta[4],theta[8],theta[12]), data %>% filter(length==50))
  nll = nll_1 + nll_10 + nll_20 + nll_50
  return(nll)
}

fit_F3 <- optim(c(0,0,0), nll_F3, method = 'BFGS', data = strength.data)
fit_F4 <- optim(c(1,1,1,1,1,1,1,1,1,1,1,1), nll_F4, method = 'BFGS', data = strength.data)
fit_F3$par
fit_F4$par
```

## Question 3

Using $\mathcal F_i$, for $i=0,1,2,3,4$ compute the standardised
residuals using $s(\boldsymbol{y})=\boldsymbol{y}$ and then plot them
against $E_{\boldsymbol{\theta}}(s(\boldsymbol{Y}))$. In the light of
the observed data $\boldsymbol{y}$, what can you say about model
$\mathcal F_i$ in each case? You can use the fact that if
$Y\sim Weibull(\eta,\tau)$ then
$$\mbox{E}[Y]=\tau\,\Gamma(1+1/\eta)\,,\qquad \mbox{V}(Y)=\tau^2\left[\Gamma(1+2/\eta)-(\Gamma(1+1/\eta))^2\right]$$
where $\eta$ is the shape parameter and $\tau$ is the scale parameter.
For the generalised Weibull you should use the function `integrate` to
integrate numerically to obtain the corresponding means and variances.

## Question 4

Consider a different type of residuals based on the empirical survival
function:
$$s(\boldsymbol{y})=\widehat{S}(t)=\frac{1}{n}\sum_{i=1}^n I(y_i>t)
$$

for each fibre length group. This is clearly an unbiased estimator of
the unknown survival function $S_*(t)=P_*(Y>t)$, that is
$E[\widehat{S}(t)]=S_*(t)$. We will define the residuals based on the
vector

$$\boldsymbol{s}=(\widehat{S}(y_{(1)}),\widehat{S}(y_{(2)}),\ldots,\widehat{S}(y_{(n)}))^T$$
so that the mean under a model with parameters $\boldsymbol{\theta}$ is
given by
$$\mu_{\boldsymbol{\theta}}=(S(y_{(1)}|\boldsymbol{\theta}),S((y_{(2)}|\boldsymbol{\theta}),\ldots,S((y_{(n)}|\boldsymbol{\theta}))^T$$
As a function of $t$ the estimator $\widehat{S}(t)$ is a decreasing step
function which jumps down sequentially at the (increasingly) ordered
observation values $y_{(1)}\leq y_{(2)}\leq \cdots\leq y_{(n)}$.

The covariance function is given by

$$Cov(\widehat{S}(t),\widehat{S}(r))=\frac{S(\max(t,r))-S(t)S(r)}{n}$$
and this defines a matrix which we call
$\boldsymbol{\Sigma}_{\boldsymbol{\theta}}$. so we can define the
residuals as

$$\hat{\epsilon}(\boldsymbol{y}):=\boldsymbol{\Sigma}^{-1/2}_{\widehat{\boldsymbol{\theta}}}(\boldsymbol{s}-\mu_{\widehat{\boldsymbol{\theta}}})$$

The function `ecdf` in base R does almost exactly what we need here
except for the fact that it uses $I(y_i\leq t)$ instead of $I(y_i>t)$.
But we can substract from one to do what we need! The output of `ecdf`
is actually a function of $t$.

Plot these residuals for each group of fibres and for each of the
models.

## Question 5

Plot the values of the attained log-likelihood at the corresponding
maximum for all the 5 families. What do you observe? Compute the Akaike
information criterion (AIC), Bayesian information criterion (BIC) and
Network information criterion (NIC) for each model and determine which
one is the best model according to each criterion.
